Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response

### 2024-12-08 paper/paper.qmd
Add information on varibles sellection and visualize selected variables

library(tidyverse)
library(dplyr)
library(lubridate)

@@

### Preview data ###
kable(head(clean_data, 6))
```

@@

## Data Cleaning
The raw dataset was stored in its original csv format to ensure data integrity and reproducibility, and later analysis data after cleaning and processing is saved into parquet file using `write_parquet` function in `arrow` package [@arrow2024].

@@

Variables removed:

Kept variables: The remaining variables are essential for the completeness of the analysis. For categorical data entries, missing values are replaced with “unreported.”

@@

## Limitations and Biases in the Data {#sec-second-point}
### Potential Bias

@@

#### Understand what is related to Severity ####

@@

# Appendix {-}

# Data and Variable Selection

```{r}
#| include: false
#| warning: false
#| message: false
#### Workspace setup ####
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(knitr)

#### Read data ####
fire_data <- read_parquet(here::here("data/02-analysis_data/tfs_analysis_data.parquet"))
#### Data Overview ####
# Structure
head(fire_data)
str(fire_data)

# Missing Value
missing_data <- colSums(is.na(fire_data))
print(missing_data)

@@

This appendix contains all plot/summary distribution of all variables in interested choose in analysis data set.

1.  Are of Origin (see @fig-area-origin-distribution)
2.  Civilian Casualties (see @tbl-civilian-casualties-summary)
3.  Estimated Dollar Loss (see @tbl-estimatedloss)

@@

4.  Final Incidence Type (see @fig-final-incident-type-distribution)
5.  Fire Alram Conditions (see @fig-firealarmsystem)

@@

6.  Ignition Source (see @fig-grouped-ignition-source-distribution)
7.  Number of Responding Apparatus (See @fig-responding-apparatus)

@@

8.  Number of Responding Personnel (See @fig-responding-personnel-hist)

@@

9.  Possible Cause (See @tbl-possible-cause-full)

@@

11. Sprinkler System Presence (@tbl-sprinkler-system)

@@

12. tfs_alarm_time (Amount of Fire Incidents by Year, see @fig-fire-incidents-by-year) (Fire Incidents by Month and Year, See @fig-year-month-incident-count)

@@

14. Firefighter Casualties (See @fig-firefighter-casulties)

@@

15. Response Time (See @fig-response-time-distribution)

>>>>

library(tidyverse)
library(dplyr)
library(lubridate)
library(kableExtra)

@@

### Preview data ###
head_table <- head(clean_data)
kable(head(head_table)) |>
  kable_styling(latex_options = c("scale_down"))

@@

## Data Cleaning

The raw dataset was stored in its original csv format to ensure data integrity and reproducibility, and later analysis data after cleaning and processing is saved into parquet file using `write_parquet` function in `arrow` package [@arrow2024].

@@

@@

## Limitations and Biases in the Data {#sec-second-point}

### Potential Bias

@@

@@

# Appendix {.unnumbered}

# Variable Selection and Visualization

```{r}
#| inlcude: false
#| warning: false
#| message: false
#| echo: false
#### Workspace setup ####
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(knitr)

# Since I want to study possible sources leading to different severity of fire, no info about sources should be dropped
clean_data <- clean_data |>
  filter(ignition_source != "999 - Undetermined")

### Ensure Datetime Consistency ###

# Convert timestamps to POSIXct format (YYYY-MM-DD HH:MM:SS)
clean_data$tfs_alarm_time <- ymd_hms(clean_data$tfs_alarm_time)  
clean_data$tfs_arrival_time <- ymd_hms(clean_data$tfs_arrival_time) 
clean_data <- clean_data[!is.na(clean_data$tfs_alarm_time), ]
clean_data <- clean_data[!is.na(clean_data$tfs_arrival_time), ]

@@

## Variables Selection Criterion

From this comprehensive dataset which included 43 vairables, I first selected the variables that are most relevant to my topic of study, leaving a total of 17 variables. Variables that are not essential to the study's focus include:

**Geographical Information of Incidence**: The study focuses on analyzing fire data for Toronto as a whole. Geographic details such as the specific station area, ward, or intersection are not critical to determining severity levels. Including these could even divert focus from more relevant variables.

-   `Latitude`, `Longitude`, `Incident_Station_Area`, `Incident_Ward`, and `Intersection`.

**Some Safety System-Related Variables**: The study aims to study some infrasture system suhc as alarm and sprinkler systems. Some variables providing redundant information about these systems were excluded. The dataset also included variables on smoke alarm systems; however, due to the excessive number of null values, these variables were excluded as they made meaningful analysis challenging.

-   `Fire_Alarm_System_Presence`, `Fire_Alarm_System_Impact_on_Evacuation`, `Smoke_Alarm_at_Fire_Origin`, `Smoke_Alarm_at_Fire_Origin_Alarm_Failure`, `Smoke_Alarm_at_Fire_Origin_Alarm_Type`, and `Sprinkler_System_Operation`.

**Repeated Information**: Certain variables contained information that overlapped with variables already chosen for inclusion. To simplify the analysis, these redundant features were removed.

-   `Count_of_Persons_Rescued`, `Estimated_Number_Of_Persons_Displaced`, `Material_First_Ignited`, `Initial_CAD_Event_Type`, `Level_Of_Origin`, `Extent_Of_Fire`, `Status_of_Fire_On_Arrival`, and `Property_Use`.

**Unnecessary Details**: There are detailes in the dataset that is not directly related to the purpose of this study. Incident-specific details, such as how long firefighters took to respond or clear the fire, are more relevant for evaluating operational efficiency. These do not directly contribute to understanding the severity levels.

-   `Ext_agent_app_or_defer_time`, `Fire_Under_Control_Time`, `Last_TFS_Unit_Clear_Time`, `Method_Of_Fire_Control`, and `Incident_Number`.

Upon reviewing the 17 initially selected variables of interest, I observed that some columns contain a significant proportion of missing values (NAs), see @tbl-missing-overview. As a result, certain columns (with more than 20% missing data) should be considered for removal if their relevance to the study topic is not as significant as other variables that should remain.

```{r}
#| label: tbl-missing-overview
#| fig-cap: This table shows the percentage of missing values for each variable of interest. Variables such as 'area_of_origin,' 'building_status,' and 'business_impact' exhibit high missing percentages, indicating potential areas for further investigation or data imputation, while others like '_id' and 'final_incident_type' have negligible missing values. Understanding these missing values is crucial for the completeness and quality of the analysis..
#| warning: false
#| message: false
#| echo: false
# Missing Value
missing_values <- colSums(is.na(clean_data))
missing_percentage <- round((missing_values / nrow(clean_data)) * 100, 2)
missing_info_table <- data.frame(Missing_Percentage = missing_percentage)
kable(missing_info_table)
```

The variables removed are:

Kept variables: These variables are essential for the completeness of the analysis. Since all of them are categorical data entries, missing values are replaced with “unreported.”

After dealing with columns with a big proportion of missing values, removing rows with a certain amount of missing values would still remain in a relatively representative dataset. Therefore, I decided to remove any columns with more than 30% missing values to the usability of the dataset fro analysis.

## Distribution and Summary of Cleaned Variables

Below are the plots or summary statistics showing the distribution of all variables of interest in the analysis dataset, after cleaning and removing columns. This includes the created variable 'Response Time.'. The description of each variable is sourced from the data portal in Open Data Toronto \[@TFSfireedata\].

1.  **Area of Origin**: OFM Area of Origin code and description, specifying where the fire started. This information will contribute to the understanding what are some origins of fire and how it may influence fire severity or outcomes. 75 different recorded origins, grouped into 11 larger groups. (see @fig-area-origin-distr)
2.  **Civilian Casualties**: Count of civilian casualties (injuries or fatalities). This is directly relates to the severity of fire incidents.(see @tbl-civilian-casualties-summary) 
3.  **Estimated Dollar Loss**: Estimated financial loss caused by the fire. This is directly relates to the severity of fire incidents. (see @tbl-estimatedloss)

@@

4.  **Final Incidence Type**: Final classification of the fire incident This can  give information about the nature of the incident, helpful to determine severity. (see @fig-final-incident-type-distribution)
5.  **Fire Alarm System Operation**: Describes the operation of fire alarm systems during the incident. This is important for assessing the role of alarms, this is a part of fire infrastructure.(see @fig-firealarmsystem)

@@

6.  **Ignition Source**: OFM Ignition Source code and description, detailing how the fire started. This is for understanding factors that contribute to fire ignition (see @fig-grouped-ignition-source-distribution)
7.  **Number of Responding Apparatus**: Number of TFS apparatus that responded to the incident. This may correlate with the severity as it reflects the scale of response. (See @fig-responding-apparatus)

@@

8.  **Number of Responding Personnel**: Number of TFS personnel that responded to the incident. This reflect the magnitude of resources deployed, could relate with fire’s severity similar as `Number_of_Responding_Apparatus`. (See @fig-responding-personnel-hist)

@@

9.  **Possible Cause**: OFM Possible Cause code and description, explaining the potential cause of the fire. This tells the circumstances leading to the fire. (See @tbl-possible-cause-full)

@@

11. **Sprinkler System Presence**: Describes whether a sprinkler system was present at the location. This provided information of the sprinkler system, which is also a part of fire infrastructure. (see @tbl-sprinkler-system)

@@

12. **TFS Alarm Time**: Time stamp of when TFS was notified of the incident. This useful when analyzing response times and their impact on severity, as well as learning about the amount of incidents across times.(Amount of Fire Incidents by Year, see @fig-fire-incidents-by-year) (Fire Incidents by Month and Year, see @fig-year-month-incident-count)

@@

14. **Firefighter Casualties**: Count of TFS firefighter casualties (injuries or fatalities). This is useful when determining the severity of fire. (See @fig-firefighter-casulties)

@@

15. **Response Time**: This is created variable. It is calculated as the difference between the `TFS_Alarm_Time` (when the fire service is notified of the incident) and the `TFS_Arrival_Time`. (See @fig-response-time-distribution)
